---
title: "HW6"
author: "Erin Reilly, Santiago Rivera Reyes, Patrick Lemis"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mid-Semester Assignment - HW5
Your task on this assignment is to work in a small group (2-3 people) to address the following questions. Only
one writeup (as a .Rmd and .pdf) needs to be submitted for each group. The problems below use several
different datasets. Your document should include code, figures, and written answers to each question.


```{r}
#Load libraries
library(tidyverse)

```

## 1. Fisheries example

The code below includes the Canadian Atlantic cod population (in tonnes) over time. Build a plot of this data and explain in writing what has happened to the population. 

ANSWER: The cod population was operating at a stable level until the early 1990s when there was a stark decrease in population size. Due to this abrupt decrease it seems as though the cod have been unable to bounce back to a stable populations, thus maintaining the low popuation throughout the early 2000s.

```{r cod_pop,echo=T,eval=T, tidy=TRUE}
cod_pop <- c(1450, 1420, 1050, 2643, 1060, 1080, 1410, 1150, 740, 175, 43, 10, 12, 15, 16, 16, 28, 30, 32, 23, 12, 19, 27)
years <- c(1983, 1984, 1985, 1986, 1987, 1988, 1989, 1990, 1991, 1992, 
1993, 1994, 1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 
2004, 2005)

typeof(cod_year)

cod_year <- data.frame(cod_pop,years)

library(tidyverse)
nrow(cod_year)
ggplot(data = cod_year, aes(x = years, y = cod_pop)) + geom_point()
```

## 2-3 Population Models

The discrete-time logistic growth model has the form N [t] + r ∗ N [t] ∗ (1 − N [t]/K). It is often used to describe population growth, such as in animal and plant populations, cancer tumor growth, and bacterial growth. 

Your task is to write a for loop that runs this model from time = 1 to time = 50. The starting value of N is 0.1. The default parameter values are r = 0.4 and K = 1. 

```{r}
#Define parameters
r <- 0.4
t <- 50
K <-1

#Create vector
N <- vector(length = length(t)) #storing population data
N[1] <- 0.1 #when t = 1

#Create loop
for(i in 1:t){
  N[i+1] <- N[i] + r * N[i] * (1-N[i]/K) #for some reason you NEED the [i+1] which I thought was exponential growth exclusive
}

```

What is the behavior (e.g., does the population go extinct? are there cycles?) with the default parameter values? 

The population increases logistically to the carrying capacity. This is reached around time point 18. 
```{r}
#Plot
plot(N) 
```


What happens if you increase r? 
 
```{r}
#Define parameters
r2 <- 0.89
t2 <- 50
K2 <-1

#Create vector
N2 <- vector(length = length(t2)) #storing population data
N2[1] <- 0.1 #when t = 1

#Create loop
for(i in 1:t2){
  N2[i+1] <- N2[i] + r2 * N2[i] * (1-N2[i]/K2) #for some reason you NEED the [i+1] which I thought was exponential growth exclusive
}
#Plot
plot(N2) 
```

Increasing the r value increases the rate at which the carrying capacity is reach. For example, an r value of 0.89 causes carrying capacity to be reached around time point 8.

Graphs are provided to support conclusions.

## 4. Measles data in the UK

On Canvas (https://mycourses.unh.edu/courses/88910/files/6917786?module_item_id=2094185), I've uploaded historical data on total measles cases and deaths for Wales and the UK. Load the data into R and build figures. Use the data to explain as much as possible about what is happening with the dynamics of measles. 

ANSWER: The dynamics of measles within Wales and the UK initially started off with high rates of infections as well as deaths. Additionally it seems like the infection rates were various in the initial year of infection, before the vaccine was released in 1963. After which the graph shows the the data gets less variable as cases and deaths decreased, evening out in the early 1990s. This could also show that the vaccine was highly effective especially over long periods of time as even the most recent data (2020) has little to no infections and deaths.

```{r}
library(ggplot2)

disease_UK <- read.table(file = 'measles_wales_and_UK.txt', row.names = NULL, header = TRUE)

view(disease_UK)

ggplot(data = disease_UK, aes(x= Year, y = Notifications)) + geom_line(aes(color = Total_Deaths))

ggplot(data = disease_UK, aes(x= Year, y = Total_Deaths)) + geom_line()

ggplot(data = disease_UK, aes(y= Notifications, x = Year)) + geom_point(aes(size = Total_Deaths)) + geom_smooth(method = 'lm') + geom_segment(aes(x=1963, y = 0, xend= 1963, yend=800000, color = 'segment'), data = disease_UK) + geom_text(aes(x=disease_UK$Year[24], y = 0, label = 'Measles Vaccine Made'), size = 2)


geom_segment(aes(x = 2005, y = 0, xend = 2005, yend = 9, colour = "segment"), data = df) + 
  geom_text(aes(x=df$Date[6], y= 10, label = "Label"), colour = "black", vjust = 1, size = 5) 
```

## 5-7

The code below implements a Wright-Fisher model with genetic drift and mutation. Use and modify the code to create plots and answer the following questions:

1. What is the effect of population size (N) on the probability of extinction of A allele?
Answer: 
As we can see on the code below, smaller populations have a higher probability of extinction of the A allele while the opposite is true for larger populations which are less susceptible to extinction as they have more stable allele frequencies. 

Code:
```{r}
N_values <- c(100, 500, 1000)  # Vary population size
p_ini <- 0.1  # Initial frequency of the A allele
u <- 0.1  # Mutation rate

max.time <- 100  # Time to run simulations

sample.population <- function(p, N, u) {
  next_pop <- vector("character", length = N)
  for (i in 1:N) {
    if (sample(x = c("A", "a"), size = 1, prob = c(p, 1 - p)) == "a") {
      # Check for mutation with probability u
      mutate <- rbinom(1, size = 1, prob = u)
      if (mutate == 1) {
        new_allele <- "A"
      } else {
        new_allele <- "a"
      }
    } else {
      new_allele <- "A"
    }
    next_pop[i] <- new_allele
  }
  return(sum(next_pop == "A") / N)  # Returns the value for p
}
results <- list()

# Question 1: Effect of population size on probability of extinction
for (N in N_values) {
  p <- numeric(max.time)
  p[1] <- p_ini
  for (t in 1:(max.time - 1)) {
    p[t + 1] <- sample.population(p[t], N, u = 0)
  }
  results[[paste("N", N, sep = "_")]] <- p
}

par(mfrow = c(1, length(N_values)))
for (i in 1:length(N_values)) {
  plot(1:max.time, results[[i]], ylim = c(0, 1), type = "l", las = 1,
       ylab = "Freq(A)", xlab = "Time", main = paste("N =", N_values[i]))
}
```

2. How does the initial frequency of the A allele affect the probability it will reach fixation?

Answer: 
According to the code below, we can see that if the initial frequency of allele A is low it has a smaller probability of reaching fixation as it starts with little representation in the population. The opposite is true when the initial frequency is high as a bigger representation in the population translates into a higher chance of fixation.

Code: 
```{r}
# Set up parameter values
N <- 1000  # Population size
p_ini_values <- seq(0.1, 0.9, by = 0.1)  # Vary initial frequency of A allele
u <- 0.1  # Mutation rate

max.time <- 100  # Time to run simulations

# Custom built function with mutation (u)
sample.population <- function(p, N, u) {
  next_pop <- vector("character", length = N)
  for (i in 1:N) {
    if (sample(x = c("A", "a"), size = 1, prob = c(p, 1 - p)) == "a") {
      # Check for mutation with probability u
      mutate <- rbinom(1, size = 1, prob = u)
      if (mutate == 1) {
        new_allele <- "A"
      } else {
        new_allele <- "a"
      }
    } else {
      new_allele <- "A"
    }
    next_pop[i] <- new_allele
  }
  return(sum(next_pop == "A") / N)  # Returns the value for p
}

# Initialize a list to store results
results <- list()
# Question 2: Effect of initial frequency on probability of fixation
for (p_ini in p_ini_values) {
  p <- numeric(max.time)
  p[1] <- p_ini
  for (t in 1:(max.time - 1)) {
    p[t + 1] <- sample.population(p[t], N, u = 0)
  }
  results[[paste("p_ini", p_ini, sep = "_")]] <- p
}


options(repr.plot.width = 10, repr.plot.height = 6)  


for (i in 1:length(p_ini_values)) {
  
  par(mfrow = c(1, 1))  
  
  par(mar = c(4, 4, 2, 2))  
  
  plot(1:max.time, results[[i]], ylim = c(0, 1), type = "l", las = 1,
       ylab = "Freq(A)", xlab = "Time", main = paste("p_ini =", p_ini_values[i]))
}
```
3. Let’s now study the combined effects of mutation and genetic
drift. Modify the above code (specifically the sample.population() function) to include a probability, u, that if a small “a” allele is chosen, it will mutate to be a big A allele. An if/else statement and or the command rbinom() might be helpful. How does this affect our findings in questions 1 and 2?

Answer: 
if mutation (u) is high, it introduces new alleles and therefore it can reduce the effect of extinction, similarly higher mutation rates can slow down the fixation process as there is more diversity in the population.


Code:

```{r}
N <- 1000  # Population size
p_ini <- 0.1  # Initial frequency of the A allele
u_values <- c(0, 0.1, 0.5)  # Vary mutation rate

max.time <- 100  # Time to run simulations

sample.population <- function(p, N, u) {
  next_pop <- vector("character", length = N)
  for (i in 1:N) {
    if (sample(x = c("A", "a"), size = 1, prob = c(p, 1 - p)) == "a") {
      # Check for mutation with probability u
      mutate <- rbinom(1, size = 1, prob = u)
      if (mutate == 1) {
        new_allele <- "A"
      } else {
        new_allele <- "a"
      }
    } else {
      new_allele <- "A"
    }
    next_pop[i] <- new_allele
  }
  return(sum(next_pop == "A") / N)  # Returns the value for p
}

results <- list()

# Question 3: Effect of mutation rate on probability of extinction and stability
for (u in u_values) {
  p <- numeric(max.time)
  p[1] <- p_ini
  for (t in 1:(max.time - 1)) {
    p[t + 1] <- sample.population(p[t], N, u)
  }
  results[[paste("u", u, sep = "_")]] <- p
}

par(mfrow = c(1, length(u_values)))
for (i in 1:length(u_values)) {
  plot(1:max.time, results[[i]], ylim = c(0, 1), type = "l", las = 1,
       ylab = "Freq(A)", xlab = "Time", main = paste("u =", u_values[i]))
}

```

