---
title: "HW6"
author: "Erin Reilly"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mid-Semester Assignment - HW5
Your task on this assignment is to work in a small group (2-3 people) to address the following questions. Only
one writeup (as a .Rmd and .pdf) needs to be submitted for each group. The problems below use several
different datasets. Your document should include code, figures, and written answers to each question.


```{r}
#Load libraries
library(tidyverse)

```

## 1 - Fisheries Example

The code below includes the Canadian Atlantic cod population (in tonnes) over time. Build a plot of this
data and explain in writing what has happened to the population.

```{r}
#Load data
cod_pop <- c(1450, 1420, 1050, 2643, 1060, 1080, 1410, 1150, 740, 175, 43, 10, 12,
15, 16, 16, 28, 30, 32, 23, 12, 19, 27)
years <- c(1983, 1984, 1985, 1986, 1987, 1988, 1989, 1990, 1991, 1992, 1993, 1994,
1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 2004, 2005)

#Build plot


```


## 2-3 Population Models

The discrete-time logistic growth model has the form N [t] + r ∗ N [t] ∗ (1 − N [t]/K). It is often used to describe population growth, such as in animal and plant populations, cancer tumor growth, and bacterial growth. 

Your task is to write a for loop that runs this model from time = 1 to time = 50. The starting value of N is 0.1. The default parameter values are r = 0.4 and K = 1. 

What is the behavior (e.g., does the population go extinct? are there cycles?) with the default parameter values? 

What happens if you increase r? 

Please provide graphs to support your conclusions.

```{r}
# Model: N[t] + r * N[t] * (1-N[t]/K)

t<-1:50
N<-0.1
r<-0.4
K<-1
output<-vector(mode = "numeric", length = length(t))

for (i in t){
  model<-N[t] + r * N[t] * (1-N[t]/K)
  print(model)
  output[i] <- model
}


```

## 4 Measles data in the UK

On Canvas (https://mycourses.unh.edu/courses/88910/files/6917786?module_item_id=2094185), I’ve up- loaded historical data on total measles cases and deaths for Wales and the UK. Load the data into R and build figures. Use the data to explain as much as possible about what is happening with the dynamics of measles.

```{r}



```

## 5-7

The code below implements a Wright-Fisher model with genetic drift and mutation. Use and modify the code to create plots and answer the following questions:

1. What is the effect of population size (N) on the probability of extinction of A allele?

2. How does the initial frequency of the A allele affect the probability it will reach fixation?

3. Let’s now study the combined effects of mutation and genetic
drift. Modify the above code (specifically the sample.population() function) to include a probability, u, that if a small “a” allele is chosen, it will mutate to be a big A allele. An if/else statement and or the command rbinom() might be helpful. How does this affect our findings in questions 1 and 2?

```{r}
# Part 1: Wright-Fisher model with genetic drift and mutation
# Created by Easton R. White
# Last edited: 4-Oct-2021
# Setting up parameter values
N = 1000 # population size
p_ini = 0.1 # the initial frequency of the A allele
u = 0.1 # mutation rate from a to A
max.time = 100 # time to run simulations
# Custom built function
sample.population <- function(p, N) {
next_pop = vector("character", length = N)
for (i in 1:N) {
new_allele = sample(x = c("A", "a"), size = 1, prob = c(p, 1 - p))
next_pop[i] = as.character(new_allele)
}
return(sum(next_pop == "A")/N) # returns the value for p
}
p = vector(mode = "numeric", length = max.time)
p[1] = p_ini
for (t in 1:(max.time - 1)) {
p[t + 1] = sample.population(p[t], N)
}
plot(1:max.time, p, ylim = c(0, 1), type = "l", las = 1, ylab = "Freq(A)", xlab = "Time")

```

