---
output:
  pdf_document: default
  html_document:
    fig.caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=TRUE)
```


# Small group assigment: Working with data in R

---

Your task on this assignment is to work in a small group (2-3 people) to address the following questions. Only one writeup (as a .Rmd and .pdf) needs to be submitted for each group. The problems below use several different datasets. Your document should include code, figures, and written answers to each question. 



## 1. Fisheries example

The code below includes the Canadian Atlantic cod population (in tonnes) over time. Build a plot of this data and explain in writing what has happened to the population. 

ANSWER: The cod population was operating at a stable level until the early 1990s when there was a stark decrease in population size. Due to this abrupt decrease it seems as though the cod have been unable to bounce back to a stable populations, thus maintaining the low popuation throughout the early 2000s.

```{r cod_pop,echo=T,eval=T, tidy=TRUE}
cod_pop <- c(1450, 1420, 1050, 2643, 1060, 1080, 1410, 1150, 740, 175, 43, 10, 12, 15, 16, 16, 28, 30, 32, 23, 12, 19, 27)
years <- c(1983, 1984, 1985, 1986, 1987, 1988, 1989, 1990, 1991, 1992, 
1993, 1994, 1995, 1996, 1997, 1998, 1999, 2000, 2001, 2002, 2003, 
2004, 2005)

typeof(cod_year)

cod_year <- data.frame(cod_pop,years)

library(tidyverse)
nrow(cod_year)
ggplot(data = cod_year, aes(x = years, y = cod_pop)) + geom_point()
```



## 2-3. Population models

The discrete-time logistic growth model has the form $N[t] + r*N[t]*(1 - N[t]/K)$. It is often used to describe population growth, such as in animal and plant populations, cancer tumor growth, and bacterial growth. Your task is to write a for loop that runs this model from time = 1 to time = 50. The starting value of N is 0.1. The default parameter values are $r=0.4$ and $K=1$. What is the behavior (e.g., does the population go extinct? are there cycles?) with the default parameter values? What happens if you increase $r$? Please provide graphs to support your conclusions.
```{r}

```


## 4. Measles data in the UK

On Canvas (https://mycourses.unh.edu/courses/88910/files/6917786?module_item_id=2094185), I've uploaded historical data on total measles cases and deaths for Wales and the UK. Load the data into R and build figures. Use the data to explain as much as possible about what is happening with the dynamics of measles. 

The dynamics of measles within Wales and the UK initially started off with high rates of infections as well as deaths. Additionally it seems like the infection rates were various in the initial year of infection, before the vaccine was released in 1963. After which the graph shows the the data gets less variable as cases and deaths decreased, evening out in the early 1990s. This could also show that the vaccine was highly effective especially over long periods of time as even the most recent data (2020) has little to no infections and deaths.

```{r}
library(ggplot2)

disease_UK <- read.table(file = 'measles_wales_and_UK.txt', row.names = NULL, header = TRUE)

view(disease_UK)

ggplot(data = disease_UK, aes(x= Year, y = Notifications)) + geom_line(aes(color = Total_Deaths))

ggplot(data = disease_UK, aes(x= Year, y = Total_Deaths)) + geom_line()

ggplot(data = disease_UK, aes(y= Notifications, x = Year)) + geom_point(aes(size = Total_Deaths)) + geom_smooth(method = 'lm') + geom_segment(aes(x=1963, y = 0, xend= 1963, yend=800000, color = 'segment'), data = disease_UK) + geom_text(aes(x=disease_UK$Year[24], y = 0, label = 'Measles Vaccine Made'), size = 2)


geom_segment(aes(x = 2005, y = 0, xend = 2005, yend = 9, colour = "segment"), data = df) + 
  geom_text(aes(x=df$Date[6], y= 10, label = "Label"), colour = "black", vjust = 1, size = 5) 
```

Measles vaccine in 1963

## 5-7. Population genetics

The code below implements a Wright-Fisher model with genetic drift and mutation. Use and modify the code to create plots and answer the following questions. 

```{r, echo=TRUE, eval=FALSE}
# Part 1: Wright-Fisher model with genetic drift and mutation

# Created by Easton R. White

# Last edited: 4-Oct-2021

# Setting up parameter values
N=1000      # population size 
p_ini= 0.1   # the initial frequency of the A allele
u = 0.1      # mutation rate from a to A
max.time=100 # time to run simulations


# Custom built function
sample.population <- function(p,N){
  next_pop = vector('character',length = N)
  for (i in 1:N){
    new_allele = sample(x = c('A','a'),size = 1,prob=c(p,1-p))
    next_pop[i] = as.character(new_allele)
  }
  return(sum(next_pop=='A')/N) # returns the value for p
}



p = vector(mode = 'numeric',length = max.time)
p[1]=p_ini

for (t in 1:(max.time-1)){
  p[t+1] = sample.population(p[t],N)
}

plot(1:max.time,p,ylim=c(0,1),type='l',las=1,ylab='Freq(A)',xlab='Time')
```


### Population genetics questions 

# 1. What is the effect of population size (N) on the probability of extinction of A allele?

# 2. How does the initial frequency of the A allele affect the probability it will reach fixation?

# 3. Let's now study the combined effects of mutation and genetic drift. Modify the above code (specifically the sample.population() function) to include a probability, u, that if a small "a" allele is chosen, it will mutate to be a big A allele. An if/else statement and or the command rbinom() might be helpful. How does this affect our findings in questions 1 and 2?




